\bf Model Visualization Techniques for a Social Network Model

### 2017-09-15 paper.Rnw
talk more about lineups

<<alldataviz, fig.cap="The smaller friendship network data we will be modelling throughout the paper. In the first wave, we can see there are two large, separate friend group. By the second wave, three students with heavier drinking behavior have separated from their group, while the others become members of the other group. By the third wave, the large group has almost completely broken up. We want to capture these changes with our SAOMs.", out.width='.99\\textwidth', fig.height=3>>=
actual1 <- merge(data.frame(as.edgelist(as.network(fd2.w1))), 
                 data.frame(id = 1:16, drink = drink2[,1]),
                 by.x = "X1", by.y = "id", all = T)
actual1$Wave <- 1 
actual2 <- merge(data.frame(as.edgelist(as.network(fd2.w2))),
                 data.frame(id = 1:16,drink = drink2[,2]),
                 by.x = "X1", by.y = "id", all = T)
actual2$Wave <- 2
actual3 <- merge(data.frame(as.edgelist(as.network(fd2.w3))),
                 data.frame(id = 1:16,drink = drink2[,3]),
                 by.x = "X1", by.y = "id", all = T)
actual3$Wave <- 3
alldat <- rbind(actual1, actual2, actual3)
#alldat$X1 <- paste0("V", alldat$X1)
alldat$X1 <- as.factor(alldat$X1)
#alldat$X2 <- ifelse(!is.na(alldat$X2), paste0("V", alldat$X2), NA)
alldat$X2 <- as.factor(alldat$X2)
#write_csv(alldat, "smallfriends4Geomnet.csv")
alldataviz <- ggplot(data = alldat, 
       aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 5, arrowsize = .25, arrowgap = .03, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 4, vjust = .5) +
  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior", labels = c("never", "1-2 times per year", "once per month", "once per week")) + 
  ThemeNet +
  facet_wrap(~Wave, labeller = "label_both") + 
  theme(legend.position = 'bottom', 
        panel.background = element_rect(fill = NA, color = "grey30"), 
        legend.text = element_text(size = 12, family = "Times New Roman"), 
        legend.title = element_text(size = 14, family = "Times New Roman"))
alldataviz
@

To guide our visual exploration of stochastic actor-oriented models, we use two data sources. The first is a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study" that is provided on the \texttt{RSiena} webpage. These data come from \citet{friendsdata}, and we chose to only work with a subset of the data to make network visualizations less busy and to make any changes in the network more noticeable. The subset contained actors 20 through 35 and the ties between them, as well as the drinking behavior of each actor at each of the three waves. This specific subset was chosen because it showed somewhat higher connectivity than other subsets, as we've emphasized in the visualizations of the three network adjacency matrices in Figure~\ref{fig:adjmatviz}. For model fitting, we condition on wave 1 and estimate the parameters of the models from the second and third waves. We will also be working with one actor level categorical covariate, drinking behavior. This variable has five values in the original data: (1) does not drink, (2) drinks once or twice a year, (3) drinks once a month, (4) drinks once a week, and (5) drinks more than once a week. The network and the actor covariate values are visualized using a node-link diagram in Figure~\ref{fig:alldataviz}.

@@

Another way to view the model in the data space is through simulation from the model. No single network alone simulated from a SAOM is going to look like the data or represent the model, just as no single value simulated from the standard normal distribution will look like a bell-shaped curve. It would therefore better to visualize many simulations together. From a statistician's point of view, a stumbling block with statistical network models generally is the lack of an ``average network" measure. Statisticians frequently rely on averages and expected values in data analyses, but statistical network models, especially those as complex as SAOMs, lack a single, intuitive expected value measure. We could talk about expected values of parameters, but the parameters can be hard to interpret. Furthermore, there is no way to talk about the expected value of an observation from a statistical network model. How then, can we arrive at an ``average" network? We propose to answer this question through visualization. For network data, one way we propose to come up with an "average" network is through a summary network drawn using the traditional node-link diagram. In Figure~\ref{fig:summnetM1}, we show an \textit{average network} created with 1,000 simulations of the second wave of the network from Model 1. \st{To make this average network, we first simulated 1,000 wave 2 and wave 3 observations of our small friendship example data from model $M1$, for which parameters had previously been estimated. We then combine the 1,000 instances of wave 2, and count up the number of times each edge appears in a simulation. Then, we combine these 1,000 networks into a single network with edgeweight equal to the proportion of time that edge appears in our 1,000 simulations. This is the network we draw using the node-link diagram.} An edge is only drawn in the average network if it appears in more than 5\% of the simulations (in at least 51 of the 1000 simulations), with edges that appear more frequently emphasized by thicker linewidths and a darker color. On either side of the average network in Figure\ref{fig:summnetM1}, we show the actual data, wave 1 on the left, and wave 2 on the right. \st{We can see that the structure of the average network is much more similar to the first network observation than to the second network observation. The simulations, however, are supposed to represent the second wave of data, which is shown on the right in Figure~\ref{fig:summnetM1}. This is an indication that our simple model, $M1$ is doing a very poor job of capturing the change mechanism from the first to the second wave of observation.} % XXX ST - I was looking at the wrong one. I had wave == 2 instead of wave == 1 in the code. This means I was looking at the wrong simulations, so they really do look more like the first data obs than the second obs. XXX Comparing the three visualizations side-by-side shows us that the summary network of simulations of wave 2 looks more like the true wave 2 than wave 1, which is a good indication that the model is capturing the true data structure. It's not ideal, however, because there appear to be too many reciprocated ties in the summary network, far more than there are in the data. This could indicate that we should remove the reciprocity parameter from the model.

<<gof2, fig.cap="An example of viewing the distribution of a statistic of interest from a SAOM. Here, the Mahalanobis distance of the cumulative outdegree values computed on a simulated network to the mean value simulated is shown. The distance computed from the data is shown as a vertical line.", fig.height=3>>=
load("data/ansnullpaper.rda")
library(lattice)
gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
qplot(x = log(gof1$`Period 1`$SimulatedTestStat), binwidth = .2) + 
  geom_vline(xintercept = log(gof1$`Period 1`$ObservedTestStat)) + 
  ThemeNoNet + 
  labs(x = "log(Statistic Value)", title = "Distribution of log Mahalanobis Distance")

@@

\st{Because longitudinal network data consist of three different ``spaces" of data, viewing the model in the data space can depend on which aspect of the model and the data you are interested in viewing. The SIENA software includes a method for modelling covariate change in the network by incorporating the network structure, assuming that the ties between nodes can affect how nodes behave over time. In this instance, a time series plot of predicted covariate values for the nodes as the network evolves is viewing the model in the time and node spaces. If, as discussed in Section~\ref{sec:modgof},
%Figure~\ref{fig:sienagof}, 
summary statistics computed on the network are of interest, distributions of these statistics computed on networks simulated from the model show the model in the edge data space. For an example, see Figure~\ref{fig:gof2}, where the distribution of the log of the Mahalanobis distance is shown for the outdegree distribution auxiliary statistics, and the log Mahalanobis distance for the observed data is shown as a vertical line.

@@

Another potential goodness-of-fit visualization is a lineup like those proposed in \citet{buja:2009}. A \texttt{lineup} ``asks the witness to identify the plot of the real data from among a set of decoys, the null plots, under the veil of ignorance" \citep[p. 4369]{buja:2009}. It can be thought of like a police lineup, where the "suspect" is in a lineup among several innocent lookalike fillers, and a witness picking the suspect out of the lineup is considered evidence against the suspect. In data and model visualization, the "suspect" is a plot of the true data, while the "filler" is composed of several plots of mock data, simulated from a hypothesized model. If the true data stands out among the simulated data, that is taken as evidence of poor model fit, whereas if the true data is difficult to identify among the simulated data, that is taked as evidence of good model fit. In Figure~\ref{fig:lineup}, the second wave of the small friendship network is shown among five simulated networks from model $M1$ using parameter values estimated from the data. The lineup allows us to view the model in the data space by placing instances of the model side-by-side with the data and examining the differences.}

@@

\subsection{Collections are more informative than singletons}

@@

To our small frienship network, we fit three models, $M1$, $M2$, and $M3$, using \texttt{RSiena} 1,000 times each. We then looked at the distribution of the fitted values, which are shown in Figure~\ref{fig:distests}. We can see from these distributions that the inclusion of the jumping transitive triplet parameter, $\beta_3$ is obviously affecting the distributions of the other four parameters included in all models, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$. In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties. It is not clear, however, that the addition of a parameter to the objective function \textit{should} effect the estimation of the rate parameters, so we continue to explore the collection of parameter estimates.

>>>>

<<alldataviz, fig.cap="The smaller friendship network data we will be modelling throughout the paper. In the first wave, we can see there are two large, separate friend group. By the second wave, three students with heavier drinking behavior have separated from their group, while the others become members of the other group. By the third wave, the large group has almost completely broken up. We want to capture these changes with our SAOMs.", out.width='.99\\textwidth', fig.height=3>>=
actual1 <- merge(data.frame(as.edgelist(as.network(fd2.w1))), 
                 data.frame(id = 1:16, drink = drink2[,1]),
                 by.x = "X1", by.y = "id", all = T)
actual1$Wave <- 1 
actual2 <- merge(data.frame(as.edgelist(as.network(fd2.w2))),
                 data.frame(id = 1:16,drink = drink2[,2]),
                 by.x = "X1", by.y = "id", all = T)
actual2$Wave <- 2
actual3 <- merge(data.frame(as.edgelist(as.network(fd2.w3))),
                 data.frame(id = 1:16,drink = drink2[,3]),
                 by.x = "X1", by.y = "id", all = T)
actual3$Wave <- 3
alldat <- rbind(actual1, actual2, actual3)
#alldat$X1 <- paste0("V", alldat$X1)
alldat$X1 <- as.factor(alldat$X1)
#alldat$X2 <- ifelse(!is.na(alldat$X2), paste0("V", alldat$X2), NA)
alldat$X2 <- as.factor(alldat$X2)
#write_csv(alldat, "smallfriends4Geomnet.csv")
alldataviz <- ggplot(data = alldat, 
       aes(from_id = X1, to_id = X2, color = as.ordered(drink), group = as.ordered(drink))) +
  geom_net(fiteach = FALSE, directed=TRUE, layout.alg = 'fruchtermanreingold', size = 5, arrowsize = .25, arrowgap = .05, ecolour = 'grey40', labelon=T, labelcolour = 'black', fontsize = 4, vjust = .5) +
  scale_color_brewer(palette = "YlOrRd", name = "Drinking\nBehavior", labels = c("never", "1-2 times per year", "once per month", "once per week")) + 
  ThemeNet +
  facet_wrap(~Wave, labeller = "label_both") + 
  theme(legend.position = 'bottom', 
        panel.background = element_rect(fill = NA, color = "grey30"), 
        legend.text = element_text(size = 12, family = "Times New Roman"), 
        legend.title = element_text(size = 14, family = "Times New Roman"))
alldataviz
@

To guide our visual exploration of stochastic actor-oriented models, we use two data sources. The first is a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study" that is provided on the \texttt{RSiena} webpage. These data come from \citet{friendsdata}, and we chose to only work with a subset of the data to make network visualizations less busy and to make any changes in the network more noticeable. To determine which subset to select, we visualized all waves of the full network using the adjacency matrix visualization approach, which we show in Figure~\ref{fig:adjmatviz}. This adjacency matrix visualization is different from the one in Figure~\ref{fig:toyplots} because it does not show the node IDs on the axes. The network in Figure~\ref{fig:adjmatviz} is much larger than our toy data example, so we remove the node labels to remove clutter. In all three adjacency matrices, the ego and the alter nodes are ordered by node ID, 1-50, which were determined arbitrarily by the relationships in wave 1. The subset we selected is outlined in red in the visualization. This subset contained actors 20 through 35 and the ties between them, as well as the drinking behavior of each actor at each of the three waves. This specific subset was chosen because it showed somewhat higher connectivity than other subsets, as we've emphasized in the visualizations of the three network observations in Figure~\ref{fig:adjmatviz}. For model fitting, we condition on wave 1 and estimate the parameters of the models from the second and third waves. We will also be working with one actor level categorical covariate, drinking behavior. This variable has five values in the original data: (1) does not drink, (2) drinks once or twice a year, (3) drinks once a month, (4) drinks once a week, and (5) drinks more than once a week. The network and the actor covariate values are visualized using a node-link diagram in Figure~\ref{fig:alldataviz}. \st{In the node-link diagram, the nodes are colored according to the drinking behavior of that student. Over time, we can see that the students tend to drink more and become increasingly isolated into smaller groups. An analysis of this type of data with a SAOM should capture these dynamics in a way that allows the researcher to draw conclusions about the nature of the network and behavioral forces at play.}

@@

Another way to view the model in the data space is through simulation from the model. No single network alone simulated from a SAOM is going to look like the data or represent the model, just as no single value simulated from the standard normal distribution will look like a bell-shaped curve. It would therefore better to visualize many simulations together. From a statistician's point of view, a stumbling block with statistical network models generally is the lack of an ``average network" measure. Statisticians frequently rely on averages and expected values in data analyses, but statistical network models, especially those as complex as SAOMs, lack a single, intuitive expected value measure. We could talk about expected values of parameters, but the parameters can be hard to interpret. Furthermore, there is no way to talk about the expected value of an observation from a statistical network model. How then, can we arrive at an ``average" network? We propose to answer this question through visualization. For network data, one way we propose to come up with an "average" network is through a summary network drawn using the traditional node-link diagram. In Figure~\ref{fig:summnetM1}, we show an \textit{average network} created with 1,000 simulations of the second wave of the network from Model 1. \st{To make this average network, we first simulated 1,000 wave 2 and wave 3 observations of our small friendship example data from model $M1$, for which parameters had previously been estimated. We then combine the 1,000 instances of wave 2, and count up the number of times each edge appears in a simulation. Then, we combine these 1,000 networks into a single network with edgeweight equal to the proportion of time that edge appears in our 1,000 simulations. This is the network we draw using the node-link diagram.} An edge is only drawn in the average network if it appears in more than 5\% of the simulations (in at least 51 of the 1000 simulations), with edges that appear more frequently emphasized by thicker linewidths and a darker color. On either side of the average network in Figure~\ref{fig:summnetM1}, we show the actual data, wave 1 on the left, and wave 2 on the right. \st{We can see that the structure of the average network is much more similar to the first wave than to the second wave. However, the simulations are supposed to represent the second wave of data, which is shown on the right in Figure~\ref{fig:summnetM1}. This is an indication that the simple model, $M1$, is doing a very poor job of capturing the change mechanism from the first to the second wave of observation.} % XXX ST - I was looking at the wrong one. I had wave == 2 instead of wave == 1 in the code. This means I was looking at the wrong simulations, so they really do look more like the first data obs than the second obs. XXX Comparing the three visualizations side-by-side shows us that the summary network of simulations of wave 2 looks more like the true wave 2 than wave 1, which is a good indication that the model is capturing the true data structure. It's not ideal, however, because there appear to be too many reciprocated ties in the summary network, far more than there are in the data. This could indicate that we should remove the reciprocity parameter from the model.

<<gof2, fig.cap="An example of viewing the distribution of a statistic of interest from a SAOM. Here, the Mahalanobis distance of the cumulative outdegree values computed on a simulated network to the mean value simulated is shown. The distance computed from the data is shown as a vertical line.", fig.height=3, out.width='.6\\textwidth'>>=
load("data/ansnullpaper.rda")
library(lattice)
gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
qplot(x = log(gof1$`Period 1`$SimulatedTestStat), binwidth = .2) + 
  geom_vline(xintercept = log(gof1$`Period 1`$ObservedTestStat)) + 
  ThemeNoNet + 
  labs(x = "log(Statistic Value)", title = "Distribution of log Mahalanobis Distance")

@@

\st{Because longitudinal network data consist of three different ``spaces" of data, viewing the model in the data space can depend on which aspect of the model and the data we are interested in viewing. %The SIENA software includes a method for modelling covariate change in the network 
Incorporating data covariates into the network structure allows us to assess whether ties between nodes are affected by how nodes behave over time. In this instance, a time series plot of predicted covariate values for the nodes as the network evolves is viewing the model in the time and node spaces. If, as discussed in Section~\ref{sec:modgof},
%Figure~\ref{fig:sienagof}, 
summary statistics computed on the network are of interest, distributions of these statistics computed on networks simulated from the model show the model in the edge data space. For an example, see Figure~\ref{fig:gof2}, where the distribution of the log of the Mahalanobis distance is shown for the outdegree distribution auxiliary statistics, and the log Mahalanobis distance for the observed data is shown as a vertical line.

@@

Another potential goodness-of-fit visualization is a lineup like those proposed in \citet{buja:2009}. A \texttt{lineup} ``asks the witness to identify the plot of the real data from among a set of decoys, the null plots, under the veil of ignorance" \citep[p. 4369]{buja:2009}. It can be thought of like a police lineup, where the ``suspect" is in a lineup among several innocent lookalike fillers, and a witness picking the suspect out of the lineup is considered evidence against the suspect. In data and model visualization, the ``suspect" is a plot of the true data, while the ``filler" is composed of several plots of mock data, simulated from a hypothesized model. If the true data stands out among the simulated data, that is taken as evidence of poor model fit, whereas if the true data is difficult to identify among the simulated data, that is taked as evidence of good model fit. \st{An application of the lineup protocal can be found in \citet{hofmann2012graphical}, where the authors examine, for instance, the differences between polar and cartesian coordinates for plotting categorical data, and density plots and box plots for determining distributional differences. They pose questions to experiment participants such as, ``Which plot is most different from the others?" for the first example and, ``In which plot is the blue group furthest to the right?" for the distributional differences. The data visualizations examined in \citet{hofmann2012graphical} are less complex than a node-link network diagram. What questions should we ask for network data visualizations? Asking participants to identify the most different plot may be difficult, as we show} in Figure~\ref{fig:lineup}. In the network lineup, the second wave of the small friendship network is shown among five simulated networks from model $M1$ using parameter values estimated from the data. What makes these plots ``different"? It seems possible to argue for any one of the six plots in Firgure~\ref{fig:lineup} as most different. So, we guide participants to look at the overall structure of the graphs to determine which has the most and least complex structure. The least complex plot, number six, has no triplets, while the most complex plot, number three, has three triplets, and in fact, plot three is the data. We have found in an experiment where all the node-link diagrams shown are simulated, one from an ``alternative" model and the rest from a ``null" model, the triangular shape of the triplets stands out to people the most. We can use the lineup protocol to help better understand SAOMs because it allows us to view the model in the data space by placing observations the model side-by-side with the data and examining the differences.}

@@

\subsection{Visualizing collections of models}

@@

To our small friendship network, we fit three models, $M1$, $M2$, and $M3$, using \texttt{RSiena} 1,000 times each. We then looked at the distribution of the fitted values, which are shown in Figure~\ref{fig:distests}. We can see from these distributions that the inclusion of the jumping transitive triplet parameter, $\beta_3$ is obviously affecting the distributions of the other four parameters included in all models, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$. In comparison with models M1 and M3, model M2 typically has higher estimates of the rate parameter, meaning that the inclusion of the covariate statistic in the model leads to higher estimates of the number of times, on average, a node gets to change its ties. It is not clear, however, that the addition of a parameter to the objective function \textit{should} effect the estimation of the rate parameters, so we continue to explore the collection of parameter estimates.
